ZpL: a p-adic precision package

### 2017-11-16 ZpL/ZpL.tex
Plan (first tentative) of the article on the package ZpL

@@

\title{Characteristic Polynomials of p-adic Matrices}

@@

\copyrightyear{2017}
\acmYear{2017}
\setcopyright{acmlicensed}
\acmConference{ISSAC '17}{July 25-28, 2017}{Kaiserslautern, Germany}
\acmPrice{15.00}
\acmDOI{10.1145/3087604.3087618}
\acmISBN{978-1-4503-5064-8/17/07}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010147.10010148.10010149.10010150</concept_id>
<concept_desc>Computing methodologies~Algebraic algorithms</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

@@

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi_M$ of 
an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
developed previously.  When $M$ is an integral matrix whose entries are 
all given at the same precision $O(p^N)$, we give a criterion (checkable 
within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
coefficient of $\chi_M$ with more accuracy than $O(p^N)$.
In general, we provide two algorithms for determining the optimal 
precision of the coefficients of $\chi_M$ and of $M$'s eigenvalues.
We provide evidence showing that classical algorithms do not reach
this optimal precision in general.
\end{abstract}

@@

The characteristic polynomial is a fundamental invariant of a matrix: its roots
give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
In fact, the best known division-free algorithm for computing determinants
over arbitrary rings \cite{kaltofen-villard:04a} does so using the characteristic polynomial.
Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
algorithms for counting points of varieties over finite fields (see
\cite{kedlaya:01a, harvey:07a, harvey:14a}).

When computing with $p$-adic matrices, entries may only be approximated
at some finite precision $O(p^N)$.  As a
consequence, in designing algorithms for such matrices one must analyze
not only the running time of the algorithm but also the accuracy of the result.

Let $M \in M_n(\Qp)$, with every entry given at precision $O(p^N)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X{-}M)$ either using recursive row expansion or various division free
algorithms \cite{seifullin:02a,kaltofen:92a}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O((n+1)!)$, $\softO(n^5)$ and $\softO(n^{3 + \omega/2})$ operations in $K$.  Second,
while the lack of division implies that the result is accurate modulo $p^N$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

A faster approach over a field is to compute the Frobenius normal form of $M$,
which is achievable in running time $\softO(n^\omega)$ \cite{keller-gehrig:85a,pernet-storjohann:07a}.
However, the use of division frequently leads to catastrophic losses of precision.
In many examples, no precision remains at the end of the calculation.

Instead, we separate the computation of the precision of $\chi_M$ from the computation
of an approximation to $\chi_M$.  Given some precision on $M$, we use the theory developed in \cite{caruso-roe-vaccon:14a}
to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
of much of this paper.  With this precision known, the actual calculation of $\chi_M$
may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
stable algorithm (see Remark \ref{rem:lift_for_optimal}).

\medskip

\noindent
{\bf Previous contributions.}
Since the description of Kedlaya's algorithm in
\cite{kedlaya:01a}, the computation of characteristic polynomials
over $p$-adic numbers
has become a crucial ingredient in many counting-points algorithms.
For example, \cite{harvey:07a, harvey:14a} use $p$-adic cohomology
and the characteristic polynomial of Frobenius to compute zeta functions
of hyperelliptic curves.

In most of these papers, the precision analysis usually
dwells on how to obtain the matrices (\textit{e.g.} of action of Frobenius)
that are involved in the point-counting schemes.
However, the computation of their characteristic polynomials 
attracts less attention: some
refer to fast algorithms (using division), while others
apply division-free algorithms.

In \cite{caruso-roe-vaccon:15a}, the authors have begun the application
of the theory of differential precision of \cite{caruso-roe-vaccon:14a}
to the stable computation of characteristic polynomials.
They describe the optimal precision
for the characteristic polynomial, but do not give practical algorithms
to compute this optimal precision.

\medskip

\noindent
{\bf The contribution of this paper.}
This paper provides a theoretical and concrete study of the propagation 
of the precision during the computation of the characteristic polynomial 
of matrix defined over the $p$-adics. Thanks to the application of the 
general framework of differential precision developed in 
\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a}, we know that the 
optimal precision of the characteristic polynomial $\chi_M$ of a matrix 
$M \in M_n(\mathbb{Q}_p)$ is controled by the adjugate $\adj(X{-}M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a short representation of
the adjugate matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
field (instead of $O(n^3)$ for a dense representation);
\item two algorithms to compute $\adj(X{-}M)$: the first one is 
division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
the second one may perform divisions and has complexity $\softO(n^\omega)$;
\end{enumerate}
and deduce from this the following applications to the question
we are interested in:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\setcounter{enumi}{2}
\item Corollary \ref{cor:prec_gain}: a simple criterion to decide 
whether $\chi_M$ is defined at precision higher than the precision of 
$M$ (when $M$ lies in $M_n(\mathbb{Z}_p)$ and all its entries are given 
at the same precision);
\item two algorithms to compute the optimal precision on each 
coefficient of the characteristic polynomial of a matrix $M \in 
M_n(\Qp)$ (whose entries may be given at different precision): the first 
one is division-free and has complexity $\softO(n^3)$ (operations in
$\Zp$) while the second 
one may perform divisions but has complexity $\softO(n^\omega)$;
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$.
\end{enumerate}
Moreover, we provide evidence showing that classical algorithms for 
computing the characteristic polynomials do not reach the optimal 
precision in many contexts.
Note that we focus on computing the precision of $\chi_M$: we do not give a stable
algorithm for computing the characteristic polynomial itself.

\medskip

\noindent
{\bf Organization of the article.}
In Section \ref{sec:theo_study}, we review the differential
theory of precision developed in \cite{caruso-roe-vaccon:14a}
and apply it to the specific case of the characteristic polynomial.
We also describe when the characteristic polynomial will have a higher
precision than the input matrix and give a compact description
of $\adj(X-M)$, the main ingredient in the differential.

In Section \ref{sec:diffHess}, we develop $\softO(n^3)$ algorithms to approximate
the Hessenberg form of $M$, and through it to find $\adj(X-M)$ and thus
find the precision of the characteristic polynomial of $M$.  In Section \ref{sec:diffFrob},
we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\adj(X-M)$.

Finally, we propose in Section \ref{sec:optjagged}
algorithms to compute the optimal coefficient-wise precision
for the characteristic polynomial.  We also give the results
of some experiments demonstrating that these methods can lead
to dramatic gains in precision over standard interval arithmetic.
We close by describing the precision associated to
eigenvalues of a matrix.

\medskip

\noindent
{\bf Notations.} 
Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. 
We will write that $f(n)=\softO (g(n))$ if there exists some
$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k).$
We will write $M$ for an $n \times n$ matrix over $K$,
and $\chi$ the characteristic polynomial map, $\chi_M \in K[X]$ for
the characteristic polynomial of $M$ and $d\chi_M$ for the differential
of $\chi$ at $M$, as a linear map from $M_n(K)$ to the space
of polynomials of degree less than $n$.
We fix an $\omega \in \R$ such that the multiplication
of two matrices over a ring is in $O(n^\omega)$ operations in the
ring. Currently, the smallest known $\omega$ is less than
$2.3728639 $ thanks to \cite{legall:14a}. 
We will denote by $I_n$ the identity matrix of rank $n$ in $M_n(K).$
When there is no ambiguity, we will drop this $I_n$
for scalar matrices, \textit{e.g.} for $\lambda \in K$
and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n - M.$
We write $\sigma_1(M), \dots, \sigma_n(M)$ for the
elementary divisors of $M$, sorted in increasing order of valuation.
Finally, given a polynomial $\chi(X) = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$,
the \emph{companion matrix} $\mathscr{C}$ associated to $\chi$ is
\begin{equation}
\label{eq:companion}
\mathscr{C} = \left( \begin{matrix}
0 & 1 & 0 & \cdots & 0 \\
0 & 0 & 1 & \ddots & \vdots \\
\vdots & \vdots & \ddots & \ddots & 0 \\
0 & 0 & \cdots & 0 & 1 \\
-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
\end{matrix} \right).
\end{equation}

\section{Theoretical study}
\label{sec:theo_study}

\subsection{The theory of p-adic precision} \label{sec:padicprec}

We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
as a foundation for our discussion of the precision for the characteristic polynomial
of a matrix.  We will be concerned with two spaces in what follows:
the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
$K_n[X]$ of monic degree $n$ polynomials over $K$.  We also define $K_{<n}[X]$ (resp. 
${\OK}_{<n}[X]$) as the space of polynomials over $K$ (resp. over $\OK$) 
of degree strictly less than $n$.

One cannot represent
matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
together with a measure of the approximation's accuracy; there are several ways to do so.
When using \emph{jagged} precision, one gives a precision on each entry,
representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
is called the \emph{absolute precision} and may depend on $(i,j)$.
When the $N_{i,j}$ are all equal, we say that the precision datum is 
\emph{flat}.

We may also model precision more flexibly by using lattices.  We 
recall that a \emph{lattice} $H$ in $M_n(K)$ is the $\OK$-span of a vector
space basis, or equivalently the image of $M_n(\OK)$ under a 
$K$-linear transformation of $M_n(K)$ \cite[\S 2.2]{caruso-roe-vaccon:14a}.
For $M \in M_n(K)$, cosets of the form $M + H$ are special
neighborhoods of $M$; we will use them for modeling precision on $M$. 
Given a family of integers $\famN = (N_{i,j})_{i,j}$ the lattice
\[
H_{\famN} = \big\{ \, M = (a_{i,j})_{i,j} \in M_n(\Qp) \,\, | \,\, \forall i,\!j\!:
\pi^{N_{i,j}} \text{ divides } a_{i,j} \, \big\}
\]
corresponds to the jagged precision introduced above.
The benefit of working with lattices is that they behave well under 
differentiable maps \cite[Lemma 3.4]{caruso-roe-vaccon:14a}, but they
require more space and time to compute.

Jagged, flat and lattice precision is defined similarly for monic polynomials, 
taking care that no precision is specified for the leading 
coefficient which must be exactly~$1$. The underlying 
vector space is now $K_{<n}[X]$ and we define a lattice in it as the
image of ${\OK}_{<n}[X]$ under some linear mapping.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map
and $d \chi_M$ for its differential at $M$.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite[Lemma 3.4]{caruso-roe-vaccon:14a}
Let $\adj(M)$ denote the adjugate\footnote{In \cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a} the adjugate was referred to as the \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
we have $\adj(M) {=} \det(M) M^{-1}$).  Recall from
\cite[Appendix B]{caruso-roe-vaccon:14a} and \cite[\S 3.3]{caruso-roe-vaccon:15a}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
\begin{array}{rcl}
d\chi_M: \quad M_n(K) & \to & K_{<n}[X] \smallskip \\
 dM & \mapsto & \tr(\adj(X{-}M) \cdot dM).
\end{array}
\end{equation}

When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may give
the precision of $\chi_M$ precisely as
\begin{equation} \label{eq:mainlem}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
The equality in \eqref{eq:mainlem} justifies our use of the phrase \emph{optimal} precision
to denote the lattice precision $d\chi_M(H)$. To make the conditions on $H$ under
which \eqref{eq:mainlem} holds more precise, let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

\begin{prop} \label{prop:mainlem}
Let $M \in M_n(K)$ and assume that $d \chi_M$ is surjective. Let:
\[
\nu = \max\left(\sum_{i=1}^{n-1} \val(\sigma_i(M)), 0\right).
\]
Then, for all integers $a, b \geq 0$ with $a + \nu < 2b$,
any lattice $H$ such that $M_n(\pi^b \OK) \subset H \subset
M_n(\pi^a \OK)$
satisfies Eq.~\eqref{eq:mainlem}
\end{prop}

\begin{proof}
Recall \cite[Definition 3.3]{caruso-roe-vaccon:15a} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
By \cite[Proposition 3.4]{caruso-roe-vaccon:15a}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
${\OK}_{<n}[X] \subset d\chi_M(M_n(\pi^\nu \OK))$.

Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
with integral coefficients, \cite[Proposition 2.2]{caruso-roe-vaccon:15a} implies
the conclusion.
\end{proof}

For further discussion of using lattices to measure precision,
details on the computation of $d\chi_M$ in \eqref{eq:dchi}, and 
examples of additional precision in $\chi_M$,
see \cite[\S\S 2.4.3, 3.2.2]{caruso:17a}.

\subsection{Several facts about the differential of \texorpdfstring{$\chi$}{x}}

We may rephrase the condition that $d\chi_M$ is surjective using more familiar invariants attached to $M$.

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi_M$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
\item \label{pt:charmin} the characteristic polynomial of $M$ is equal to the minimal polynomial of $M$.
\end{enumerate}
\end{prop}

\begin{proof}
The equivalence of \eqref{pt:cyc}, \eqref{pt:1d}, and \eqref{pt:charmin} is standard; see \emph{e.g.}
\cite[\S 7.1]{hoffman-kunze:LinearAlgebra}.  We now show
\eqref{pt:cyc} $\Rightarrow$ \eqref{pt:surj} and \eqref{pt:surj} $\Rightarrow$ \eqref{pt:1d}

For any $P \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the last column of
$\adj(X{-}M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X{-}M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
Since $\lambda$ occurs in two Jordan blocks,
$X - \lambda$ will divide every entry of $\adj(X{-}M)$ and $d\chi_M$ will not be surjective.
\end{proof}

We also have an integral analogue of Proposition \ref{prop:surjectivity}.

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is 
${\OK}_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} is as in Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
\end{proof}

The relationship between precision and the images of lattices under $d\chi_M$ allows us to
apply Proposition \ref{prop:intsurj} to determine when the precision of the characteristic polynomial
is the minimum possible.

\begin{cor} \label{cor:prec_gain}
Suppose that $M \in \GL_n(\OK)$ is known with flat precision $O(\pi^N)$.
Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^N)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field \cite{storjohann:01a}.

\medskip

\noindent
{\bf Stability under multiplication by $X$.}
By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
often be convenient to view $d \chi_M$ as a $K$-linear mapping
$M_n(K) \to K[X]/\chi_M(X)$.

\begin{prop}
Let $R$ be the subring of $K[X]$ consisting of polynomials $f$ for
which $f(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
multiplication by $R$.
\end{prop}
\begin{proof}
Let $A = \adj(X{-}M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
with its adjugate is the determinant, $(X{-}M) \cdot A = \chi_M$ and thus
$f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $f(X) \cdot \Lambda$, while the span of the entries
of the right hand side is contained within $\Lambda$ since $f(M) \in M_n(\OK)$.
\end{proof}

\begin{cor}
If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable under multiplication by $X$, so is a module over $\OK[X]$.
\end{cor}

\medskip

\noindent
{\bf Compact form of $d \chi_M$.}
Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$ as in \eqref{eq:companion}.
Assume one of the equivalent conditions of Proposition~\ref{prop:surjectivity}, giving a matrix 
$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Then Proposition~\ref{prop:surjectivity}
also holds for $M^T$, yielding an
invertible matrix $Q \in \GL_n(K)$ with $M^T = Q \mathscr{C} Q^{-1}$.

\begin{prop}
\label{prop:shortcom}
We keep the previous notations and assumptions.
Let $Y$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
\begin{equation}
\label{eq:shortcom}
\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
\mod \chi_M
\end{equation}
for some $\alpha \in K[X]$.
\end{prop}

\begin{proof}
Write $A = \adj(X{-}M)$. From $(X{-}M) \cdot A \equiv 0 
\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} A \equiv 0 \pmod{\chi_M}$. 
Therefore each column of $P^{-1} A$ lies in the right kernel of $X{-}\mathscr{C}$
modulo $\chi_M$. On the other hand, a direct computation shows that
every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
$\chi_M$ can be written as $W = w \cdot Y^T$ for some $w \in 
K[X]/\chi_M$. We deduce that $A \equiv P \cdot Y^T B \pmod{\chi_M}$
for some row vector $B$.
Applying the same reasoning with $M^T$, we find that $B$ can be
written $B = \alpha Y Q^T$ for some $\alpha \in K[X]/\chi_M$ and
we are done.
\end{proof}

Proposition~\ref{prop:shortcom} shows that $\adj(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$: the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ entries to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Section \ref{sec:diffFrob} that interesting
information can be read off of this short form $(\alpha, P, Q, 
\chi_M)$.

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \adj(X-M) U^{-1}$ using the dense
representation, which requires $\softO(n^{\omega+1})$ operations in $K.$
\end{rem}

\section{Hessenberg form}
\label{sec:diffHess}

In this section, we combine the computation of a Hessenberg form of a 
matrix and the computation of the inverse through the Smith normal form 
(SNF) over a complete discrete valuation field (CDVF) to compute 
$\adj(X{-}M)$ and $d \chi$. If $M \in M_n(\OK)$, then only division 
by invertible elements of $\OK$ will occur.

\begin{rem}
In what follows, we will count operations in $K$ (regardless to 
precision) for expressing our complexities. This choice makes sense 
because we will not usually need much precision on the entries of 
$\adj(X{-}M)$ (roughly only their valuations matter).
\end{rem}

\subsection{Hessenberg form}

We begin with the computation of
an approximate Hessenberg form.

\begin{deftn}
A \emph{Hessenberg matrix} is a matrix $M \in M_n(K)$ with
\[
M_{i,j}=0 \mbox{ for $j \le i-2$.}
\]
Given integers $n_{i,j}$, an \emph{approximate Hessenberg matrix}
is a matrix $M \in M_n(K)$ with
\[
M_{i,j} = O(\pi^{n_{i,j}}) \mbox{ for $j \le i-2$.}
\]
If $M \in M_n (K)$ and $H \in M_n (K)$ is an (approximate) Hessenberg matrix
similar to $M$, we say that H is an \emph{(approximate) Hessenberg form} of $M.$
\end{deftn}

It is not hard to prove that every matrix over a field admits
a Hessenberg form.
We prove here that we may compute an approximate Hessenberg form 
for any matrix known at finite jagged precision over $K$.
Moreover, we provide an exact change of basis matrix.

\noindent\hrulefill

\noindent {\bf Algorithm 1:} {\tt Approximate Hessenberg form computation}

\noindent{\bf Input:} a matrix $M$ in $M_n(K).$

\smallskip

\noindent 0.\ $P:=I_n.$ \: $H:=M.$

\noindent 1.\ {\bf for} $j=1,\dots,n-1$ {\bf do}

\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $\val(H_{i_{min},j})$ is minimal.

\noindent 3.\  \:  {\bf for} $i=j+2,\dots,n$ {\bf do}

\noindent 4. \ \: \:  \textbf{Eliminate} the significant digits of $H_{i,j}$ by pivoting with row $j+1$ 
using a matrix $T.$

\noindent 5. \ \: \:  $H:=H \times T^{-1}.$ \: $P:=T \times P.$

\noindent 6. \textbf{Return} $H,P.$

\vspace{-1ex}\noindent\hrulefill

\medskip

Note that we use the notation of matrix multiplication in line 5, but actual updating of
$H$ and $P$ uses row and column operations (individually using $O(n)$ operations in $K$).

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M$.
$P$ is exact  
and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
\end{prop}
\begin{proof}
Inside the nested \textbf{for} loop, to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
the corresponding coefficient of the corresponding shear matrix is 
any exact element of $K$ congruent to 
$\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}$.
Exactness follows directly. 
The rest is clear.
\end{proof}

\begin{rem} \label{rem:char_pol_from_hessenberg}
From a Hessenberg form of $M,$ it is well known
that one can compute the characteristic polynomial of 
$M$ in $O(n^3)$ operations in $K$ \cite[pp. 55--56]{Cohen:2013}
However, this computation involves division, and its
precision behavior is not easy to quantify.
\end{rem}

\subsection{Computation of the inverse}

In this section, we prove that to compute the inverse of
a matrix over $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
We first recall the differential of matrix inversion.

\begin{lem}
Let $u \: : \: GL_n (K) \rightarrow GL_n(K),$ $M \mapsto M^{-1}.$
Then for $M \in GL_n (K),$ $du_M(dM)=M^{-1} dM M^{-1}.$
It is always surjective.
\end{lem}

We then have the following result about the loss in precision when computing the inverse.

\begin{prop}
Let $\cond(M) = \val(\sigma_n(M))$.
If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal,
at least when $N$ is large.
\end{prop}

\begin{proof}
The smallest valuation of an entry of $M^{-1}$ is $-\cond(M)$.
Consequently, $N - 2\cond(M)$
can be obtained as the valuation of an entry of $du_M(dM)$,
and it is the smallest that can be achieved this way for $dM$ in a flat precision lattice.
Hence the optimality of the bound given for large 
$N$ \cite[Lemma 3.4]{caruso-roe-vaccon:14a}.

Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
known at precision at least $O(\pi^{N-\cond(M)})$, with coefficients in $\OK$
and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{N-2\cond(M)}),$
which concludes the proof.
\end{proof}

\subsection{The adjugate of \texorpdfstring{$X{-}H$}{X-H}}

In this section, we compute $\adj(X-H)$ for a Hessenberg matrix $H$
using the Smith normal form computation of the previous section.
The entries of $\adj(X-H)$ lie in $K[X]$, which is not a complete
discrete valuation field,
so we may not directly apply the methods of the previous section.
However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
entries lie in $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
using an SNF method, with no division in $K$.

We begin by relating adjugates of
similar matrices:

\begin{lem} \label{lem:adjugate_of_similar}
If $M_1,M_2 \in M_n(K)$ and $P \in GL_n (K)$ are such that
$M_1=PM_2P^{-1},$ then:
\[ \adj (X-M_1)=P \adj (X-M_2) P^{-1}. \] 
\end{lem}

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $f \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $d$ of $f$ as $f^{\rec,d}=X^d f \left( 1/X \right).$
Let $F \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $F^{\rec,d}$ the matrix with $(F^{\rec,d})_{i,j} = (F_{i,j})^{\rec,d}$.
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{align*}
\adj(1-XM)^{\rec,n-1}&=\adj(X-M), \\
(\chi_M I_n)^{\rec,n}&=(1-XM) \adj(1-XM).\\
\end{align*}
\end{lem}
\begin{proof}
For any matrix $F \in M_d(K[X])$ of polynomials of degree at most $1$,
we have $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
This result directly implies the second part of the lemma; the first part follows
from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
are determinants of size $n-1$.
\end{proof}

This lemma allows us to compute $\adj(1-XM)$ instead of $\adj(X-M).$
This has a remarkable advantage: the pivots during the computation of
the SNF of $\adj(1-XM)$ are units in $K\llb X\rrb$ (in $\OK\llb X\rrb$ if $M \in M_n(\OK)$) and are known
in advance to be on the diagonal. This leads to a very smooth
precision and complexity behavior when $M \in M_n(\OK).$

\noindent\hrulefill

\noindent {\bf Algorithm 2:} {\tt Approximate $\adj (X -H)$ }

\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(\OK).$

\smallskip

\noindent 0.\ $U:=1-XH.$ $U_0:=1-XH.$

\noindent 1.\ While updating $U$, \textbf{track} $P$ and $Q$ so that $U_0=PUQ$ is always satisfied.

\noindent 2.\ {\bf for} $i=1,\dots,n-1$ {\bf do}

\noindent 3.\  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,j},$ for $j\geq i+1$ 
using the invertible pivot
$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \OK[X]$).

\noindent 4.\    {\bf for} $i=1,\dots,n-1$ {\bf do}

\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i+1,i},$
using the invertible pivot $U_{i,i}.$

\noindent 6. \ $\psi:=\prod_i U_{i,i}.$

\noindent 7. \ Rescale to get $U = I_n \mod X^{n+1}.$

\noindent 8. \ $Y:=\psi \times P \times Q   \mod X^{n+1}.$ \footnote{The product $P \times Q$
should be implemented by sequential row operations corresponding to the eliminations in Step 5
in order to avoid a product of two matrices in $M_n(\OK[X])$.}

\noindent 9. \ \textbf{Return} $Y^{\rec,n-1}, \psi^{\rec,n}.$

\vspace{-1ex}\noindent\hrulefill

\medskip

>>>>

\newcommand{\ZpL}{\texttt{ZpL}}
\newcommand{\ZpLCA}{\texttt{ZpLCA}}
\newcommand{\ZpLCR}{\texttt{ZpLCR}}
\newcommand{\ZpLF}{\texttt{ZpLF}}

@@

\title{ZpL: a p-adic precision package}

@@

@@

\begin{abstract}
\end{abstract}

@@

\begin{itemize}
\item Interval arithmetic, floating point arithmetic
\item short introduction of the precision lemma and lattices
\item goal: tracking precision using this theory
\item short presentation of the package
\item plan of the article
\end{itemize}

\section{Examples}

\begin{itemize}
\item Elementary arithmetic
\item Determinant + matrix multiplication
\item gcd of polynomials
\item some benchmarks on precision
\end{itemize}

\section{Theory}

\subsection{The precision Lemma}

\subsection{The way the precision is tracked with \ZpL}

We follow the precision lattice.
Closely related to automatic differentiation.

\begin{itemize}
\item \ZpLCR, \ZpLCA: we use a cap to ensure that we have a lattice at
each step.
\item \ZpLF: we work with $\Zp$-submodules of possible positive codimension
and represent them using matrices of floating point $p$-adics.
\end{itemize}

\subsection{The notion of working precision}

\begin{itemize}
\item what is it?
\item how can we compute it? (Answer: Hermite reduction of the transpose)
\item how can we estimate it?
\end{itemize}

\subsection{Complexity}

\begin{tabular}{l|c|c|}
& \ZpLCA/\ZpLCR & \ZpLF \\
\hline
Creation of a variable & $O(n)$ & $O(n_0)$ \\
\hline
Deletion of a variable & $O(m^2)$ & $-$ \\
\hline
\end{tabular}

discussion on temporal locality

\section{Conclusion}

\begin{itemize}
\item In which case should we use this package? Mainly for exploration
\item example of $p$-adic differential equations (especially for $p=2$)
\end{itemize}
